---
title: "Some helpful tips for formatting tables using knitr::kable"
output: html_document
author: Plamena Panayotova
date: "2025-03-24"
---
<br>
<br>

The following document is for teaching purposes only. It aims to show students learning R and quantitative analysis how to format tables, charts, test and model outputs using the kable function from the knitr package. 

<br>
<br>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)


library(tidyverse)
library(ggplot2)
library(forcats)
library(gridExtra)
library(flextable)
library(knitr)
library(kableExtra)
library(flextable)
library(broom)
library(texreg)
library(car)
library(lmtest)
library(vtable)
library(sandwich)
library(skimr)



listings <- read_csv("listings_edinburgh.csv")



listings <- filter(listings,
                   price >= 25, price <= 1000, bathrooms <= 5) %>%
  drop_na(price, room_type)



listings <- mutate(listings,
                   room_type = factor(room_type),
                   property_type = factor(property_type))

```



```{r}

listings <- listings %>% 
  mutate(property_type = fct_collapse(property_type,
                                      Bespoke = c("Aparthotel", "Barn", "Boat", "Casa particular (Cuba)", "Earth house", "Hut", "Loft", "Tiny house", "Windmill", "Other" ),
                                      Apartment = c("Apartment", "Bed and breakfast", "Guest suite", "Serviced apartment", "Chalet", "Condominium"),
                                      House = c("Bungalow", "Cottage", "Guesthouse", "House", "Townhouse", "Villa"),
                                      Hotel = c("Boutique hotel", "Hotel"),
                                      Rustic = c("Cabin", "Camper/RV", "Campsite", "Farm stay", "Hostel", "Tent")
                                       )
      )


listings$property_type <- factor(listings$property_type, 
                                 levels = c("Apartment", "House", "Rustic", "Hotel", "Bespoke"))

```

# 1. Descriptive statistics

<br>


## 1.1 Variables:

```{r}
table_1 <- data.frame(
  Variable = c("price", "accommodates", "room_type", "bedrooms", "bathrooms","review_scores_ratings_rating", "property_type"),
  Description = c("Price of property ($)", "Accommodation capacity", "Room type", "Number of bedrooms", "Number of bathrooms", "Review score rating", "Type of property"),
  Unit= c("US dollars", "n guests", "Entire home/apt, Private room, Shared room", "n bedrooms", "n bathrooms", "Score rating (0-100)", "Bespoke, Apartment, House,    Hotel, Rustic"))

kable(table_1, caption = "Table 1. Variable Names and Definitions. Source: InsideAirbnb.com", format = "html") %>%
  kable_styling(full_width = FALSE, position = "left")
```

<br>
<br>

## 1.2 Summary statistics and a chart of the dependent variable:


```{r}

# Table of price:

table_2 <- listings %>%
  reframe(
    Variable = "Price of property (\\$)", # the two back slashes escape the dollar sign, so the document knits properly
    Count = round(length(price), 0), 
    Mean = round(mean(price, na.rm = TRUE), 1), 
    SD = round(sd(price, na.rm = TRUE), 1), 
    Min = round(min(price, na.rm = TRUE), 1), 
    Q1 = round(quantile(price, 0.25, na.rm = TRUE), 1), 
    Median = round(median(price, na.rm = TRUE), 1), 
    Q3 = round(quantile(price, 0.75, na.rm = TRUE), 1), 
    Max= round(max(price, na.rm = TRUE), 1), 
    IQR = round(IQR(price, na.rm = TRUE), 1))



kable(table_2, caption = "Table 2. Summary Statistics of Price of Property ($). Source: InsideAirbnb.com", format = "html") %>%
  kable_styling(full_width= TRUE, position ="left")


# The below code creates a table with alternative formatting (un-comment if you want to try, I like the above version better):
# price <- listings$price
# 
# table_2.1 <- tibble(
#   Statistic = c("Count", "Mean", "Standard Deviation", "Min", "Q1", "Median", "Q3", "Max", "IQR"),
#   Value = c(
#     round(length(price), 0),  # Round count to 0 decimal places
#     round(mean(price, na.rm = TRUE), 1), 
#     round(sd(price, na.rm = TRUE), 1), 
#     round(min(price, na.rm = TRUE), 1), 
#     round(quantile(price, 0.25, na.rm = TRUE), 1), 
#     round(median(price, na.rm = TRUE), 1), 
#     round(quantile(price, 0.75, na.rm = TRUE), 1), 
#     round(max(price, na.rm = TRUE), 1), 
#     round(IQR(price, na.rm = TRUE), 1))
# )
# 
# 
# kable(table_2.1, caption = "Table 2. Summary Statistics for Price. Source: InsideAirbnb.com", format = "html") %>%
#   kable_styling(full_width= TRUE, position ="left")




```

<br>
<br>

```{r}
# A histogram of price:
listings %>%
  ggplot(aes(price)) +
  geom_histogram(colour="deepskyblue4", fill="deepskyblue3", binwidth=20) +
  labs(title = "Figure1. Distribution of price of Airbnb listings in Edinburgh",
       subtitle = "Prices restricted to range $25 to $1,000",
       x = "Price ($)",
       y = "Count",
       caption = "Source: InsideAirbnb.com") +
  scale_x_continuous(breaks = seq(0, 1000, by = 200)) +
  theme_minimal()

```

<br>
<br>

## 1.3 Summary statistics and a chart of the key independent variable:


```{r}

# Table of room_type:

table_3 <- listings %>%                       
  count(room_type) %>% 
  mutate(
    Percent = round(n / sum(n) * 100, 1)) %>%
  add_row(
    room_type = "Total",               # Label for the summary row
    n = sum(.$n),                      # Total count
    Percent = 100) %>%  
  arrange(room_type) 




kable(table_3, col.names = c("Room Type", "Count", "Percent"),
      caption = "Table 3. Frequency Table of Room Type. Source: Inside AirBnB.", format= "html") %>%
  kable_styling(full_width= TRUE, position ="left")

```

<br>
<br>
```{r}

# Bar chart of room_type:

listings %>% 
  ggplot(aes(room_type)) +
  geom_bar(colour="deepskyblue4", fill="deepskyblue3") +
  labs(title = "Figure 2. AirBnB room types in Edinburgh.",
       x = "Type of room",
       y = "Count",
       caption = "Source: InsideAirbnb.com") +
  theme_minimal()

```

<br>
<br>

## 1.4 Summary statistics of all variables:


```{r}

# Table of all numeric variables:

table_4 <- listings %>%
  reframe(
    Variable = c("Price", "Review Score", "Bedrooms", "Bathrooms", "Accommodation Capacity"),
    Count = c(sum(!is.na(price)), sum(!is.na(review_scores_rating)), sum(!is.na(bedrooms)), sum(!is.na(bathrooms)), sum(!is.na(accommodates))),
    Mean = round(c(mean(price, na.rm = TRUE), mean(review_scores_rating, na.rm = TRUE), mean(bedrooms, na.rm = TRUE), mean(bathrooms, na.rm = TRUE), mean(accommodates, na.rm = TRUE)), 1),
    SD = round(c(sd(price, na.rm = TRUE), sd(review_scores_rating, na.rm = TRUE), sd(bedrooms, na.rm = TRUE), sd(bathrooms, na.rm = TRUE), sd(accommodates, na.rm = TRUE)), 1),
    Min = c(min(price, na.rm = TRUE), min(review_scores_rating, na.rm = TRUE), min(bedrooms, na.rm = TRUE), min(bathrooms, na.rm = TRUE), min(accommodates, na.rm = TRUE)),
    Q1 = c(quantile(price, 0.25, na.rm = TRUE), quantile(review_scores_rating, 0.25, na.rm = TRUE), quantile(bedrooms, 0.25, na.rm = TRUE), quantile(bathrooms, 0.25, na.rm = TRUE), quantile(accommodates, 0.25, na.rm = TRUE)),
    Median = c(median(price, na.rm = TRUE), median(review_scores_rating, na.rm = TRUE), median(bedrooms, na.rm = TRUE), median(bathrooms, na.rm = TRUE), median(accommodates, na.rm = TRUE)),
    Q3 = c(quantile(price, 0.75, na.rm = TRUE), quantile(review_scores_rating, 0.75, na.rm = TRUE), quantile(bedrooms, 0.75, na.rm = TRUE), quantile(bathrooms, 0.75, na.rm = TRUE), quantile(accommodates, 0.75, na.rm = TRUE)),
    Max = c(max(price, na.rm = TRUE), max(review_scores_rating, na.rm = TRUE), max(bedrooms, na.rm = TRUE), max(bathrooms, na.rm = TRUE), max(accommodates, na.rm = TRUE)),
    IQR = round(c(IQR(price, na.rm = TRUE), IQR(review_scores_rating, na.rm = TRUE), IQR(bedrooms, na.rm = TRUE), IQR(bathrooms, na.rm = TRUE), IQR(accommodates, na.rm = TRUE)), 1)
  )


kable(table_4, caption = "Table 4. Summary Statistics of Key Variables. Source: Inside AirBnB.", format = "html") %>%
  kable_styling(full_width = TRUE, position = "left")
```


```{r}

# Tables of the factor variables:

table_5 <- listings %>%                       
  count(property_type) %>%
  mutate(
    Percent = round(n / sum(n) * 100, 1)
  ) %>%
  add_row(
    property_type = "Total",               # Label for the summary row
    n = sum(.$n),                          # Total count
    Percent = 100                          # Total percentage
  ) 


kable(table_5, col.names = c("Property Type", "Count", "Percent"),
      caption = "Table 5. Frequency Table of Property Type. Source: Inside AirBnB.", format= "html") %>%
  kable_styling(full_width= TRUE, position ="left")


table_6 <- listings %>%                       
  count(room_type) %>%
  mutate(
    Percent = round(n / sum(n) * 100, 1)
  ) %>%
  add_row(
    room_type = "Total",                   # Label for the summary row
    n = sum(.$n),                          # Total count
    Percent = 100                          # Total percentage
  ) 

 

kable(table_6, col.names = c("Room Type", "Count", "Percent"),
      caption = "Table 6. Frequency Table of Room Type. Source: Inside AirBnB.", format= "html") %>%
  kable_styling(full_width= TRUE, position ="left")

```


<br>
<br>

## 1.5 Summary statistics of price by room and property type:



```{r results ="asis"}

# Table of price by room type:

table_7 <- listings %>%
  group_by(room_type) %>%
  summarise(
    "Count" = n(),
    "Percent" = round((n() / nrow(listings)) * 100, 1),
    "Mean (price)" = round(mean(price), 1),
    "SD (price)" = round(sd(price), 1),
    "Median (price)" = median(price, na.rm = TRUE),
    "IQR(price)" = IQR(price, na.rm = TRUE)
  ) %>%
  rename("Type of Room" = room_type) %>%  # Rename column before printing
  kable(caption = "Table 7. Property Price ($) by Room Type. Source: Inside AirBnB.", format= "html") %>%
    kable_styling(full_width = TRUE, position = "left")


table_7

```




```{r}

# Table of price by property type:

table_8 <- listings %>%
  group_by(property_type) %>%
  summarise(
    "Count" = n(),
    "Percent" = round((n() / nrow(listings)) * 100, 1),
    "Mean (price)" = round(mean(price), 1),
    "SD (price)" = round(sd(price), 1),
    "Median (price)" = median(price, na.rm = TRUE),
    "IQR(price)" = IQR(price, na.rm = TRUE)
  ) %>%
  arrange(desc(Count)) %>%
  rename("Type of Property" = property_type) %>%  # Rename column before printing
  kable(caption = "Table 8. Property Price ($) by Property Type. Source: Inside AirBnB.", format= "html") %>%
    kable_styling(full_width = TRUE, position = "left")

table_8

```



```{r}

#A less flexible alternative (uncomment if you want to try):

# table_9 <- listings %>%
#   rename(
#     "Number of Bedrooms" = bedrooms,
#     "Number of Bathrooms" = bathrooms,
#     "Accommodation Capacity" = accommodates,
#     "Price" = price,
#     "Review Score Rating" = review_scores_rating
#   ) %>%
#   sumtable(vars = c("Price", "Review Score Rating", "Number of Bedrooms", "Number of Bathrooms", "Accommodation Capacity"))
# 
# table_9


# table_10 <- listings %>%
#   rename(
#     "Type of Room" = room_type,
#     "Type of Property" = property_type) %>%
#  sumtable(vars = c("Type of Room", "Type of Property"))
# 
# table_10

```
<br>
<br>

# 2. Initial tests:
<br>
<br>

## 2.1 An ANOVA test:

```{r}
#Save the model and model summary:
anova_model_1 <- listings %>%
  aov(log(price) ~ room_type, data = .)  # Store the anova model first

anova_test_1 <- summary(anova_model_1)



tidy_anova <- tidy(anova_model_1)
tidy_anova$term[tidy_anova$term == "room_type"] <- "Type of Room"



tidy_anova <- tidy(anova_model_1) %>%
  rename(
    "Variable" = term,               # Rename 'term' to 'Variable'
    "Degrees of Freedom" = df,
    "Sum of Squares" = sumsq,
    "Mean Square" = meansq,
    "F-Value" = statistic,
    "p-Value" = p.value
  ) %>%
  mutate(Variable = ifelse(Variable == "room_type", "Type of Room", Variable),
         `p-Value` = format.pval(`p-Value`, eps = .001, digits = 2)
         )

kable(tidy_anova, caption = "Table 11. ANOVA Test Results for Room Type vs. Log Price. Source: Inside AirBnB.", format = "html") %>%
  kable_styling(full_width = TRUE, position = "left")


```
<br>
<br>

## 2.2. A t-test:

```{r}
t_test_result <-  listings %>% 
  filter(room_type != "Shared room") %>% t.test(log(price) ~ room_type, .)

t_test_table <- tidy(t_test_result)

t_test_table <- tidy(t_test_result) %>%
  select(-method, -alternative) %>%  # Remove unwanted columns
  rename(
    "Estimate (Difference)" = estimate,
    "Mean Group 1" = estimate1,
    "Mean Group 2" = estimate2,
    "t-statistic" = statistic,
    "p-value" = p.value,
    "Lower CI (95%)" = conf.low,
    "Upper CI (95%)" = conf.high
  ) %>%
  mutate(
    `p-value` = format.pval(`p-value`, eps = 0.001, digits = 3),
    across(where(is.numeric), ~ round(., digits = 3))
  )

kable(t_test_table, caption = "Table 12. t-test results for Log Price vs. Room_type. Source: Inside AirBnB.", format = "html") %>%
  kable_styling(full_width = TRUE, position = "left")

```
<br>
<br>


## 2.3 Pearson's correlation test:

```{r warning=FALSE}


cor_result <- cor.test(~ log(price) + review_scores_rating, data = listings)


cor_table <- tidy(cor_result) %>%
  mutate(
    # Rename 'review_scores_rating' to 'Review Score'
    term = "Review Score",
    # Format p-value in scientific notation if < 0.001
    p.value = format.pval(p.value, eps = 0.001, digits = 3)
  ) %>%
 rename(
    "Variable" = term,
    "Correlation (r)" = estimate,
    "t-statistic" = statistic,
    "p-value" = p.value,
    "Lower CI (95%)" = conf.low,  
    "Upper CI (95%)" = conf.high
  ) %>%
  select(
    Variable, 
    `Correlation (r)`, 
    `t-statistic`, 
    `Lower CI (95%)`,  
    `Upper CI (95%)`, 
    `p-value`
  ) %>%
  mutate(across(where(is.numeric), round, digits = 3))


kable(cor_table, caption = "Table 13. Pearson's correlation Log Price vs. Review Score. Source: Inside AirBnB.", format = "html") %>%
  kable_styling(full_width = TRUE, position = "left")

```
<br>
<br>


# 3. The models:


```{r}

m4_log_1 <- listings %>%
  mutate(log_price = log(price)) %>%
  lm(log_price ~ room_type + property_type + bathrooms + bedrooms + accommodates + review_scores_rating, 
     data = .)


```


```{r}

m4_log_2 <- listings %>%
  mutate(log_price = log(price)) %>%
  lm(log_price ~ room_type + property_type + bathrooms + bedrooms + accommodates, 
     data = .)

```


```{r results="asis"}

htmlreg(list(m4_log_1, m4_log_2),
        stars = 0.05,
          digits = 2, 
          caption = "Table 12. Regression results. Dependent variable: log(price).",
          caption.above = TRUE,
          groups = list("Room Type" = 2:3, "Property Type" = 4:7),
          custom.coef.names = c("(Intercept)","Private Room","Shared Room","Bespoke","Hotel","House","Rustic","Number of Bathrooms","Number of Bedrooms","Accomodation","Review Score"))

```


